---
title: "Women's Data"
author: "Julie Hunter"
date: "2026-08-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(lubridate)
library(coda)
```

Loading in the data and extracting a smaller set of athletes
```{r}
# Female 100m Freestyle Swimming Data
f.swim.full.data <- read.csv("Swimming_Dist_100m_Female_Freestyle.csv", header = TRUE)

# Extract the relevant columns 
f.swim.data.1 <- dplyr::select(f.swim.full.data,
                      PersonId = PersonId.x,
                      TimeSeconds,
                      ResultDate,
                      Age = AthleteResultAge,
                      PoolConfiguration,
                      PhaseName,
                      EventTypeName)

# Find the top 100 athletes (with sufficient amount of data)
  # First filter out the athletes with too few races (<15)
f.valid <- f.swim.data.1 %>%
              group_by(PersonId) %>%
              summarise(n_obs = n()) %>%
              filter(n_obs >= 15)

  # Then find the top 100 athletes 
f.pb.times <- f.swim.data.1 %>% 
                group_by(PersonId) %>% # First find personal best times for each athlete
                summarise(pb = min(TimeSeconds, na.rm = TRUE)) %>% 
                inner_join(f.valid, by = "PersonId") %>%
                arrange(pb) %>% # Then sort the pb times (fastest/best at the beginning)
                slice_head(n = 100) # Then keep only the top 100 

# Extract the data for the top 100 athletes
  # First find the PersonId for the top 100 athletes
f.top.100 <- f.pb.times$PersonId

  # Then keep only the data for these top 100 athletes
f.swim.data <- f.swim.data.1 %>%
                  filter(PersonId %in% f.top.100)

# Check to make sure each athlete has a sufficient amount of races
f.counts <- f.swim.data %>%
  group_by(PersonId) %>%
  summarise(n_obs = n())


# Change the PoolConfiguration variable to be a binary regressor
f.swim.data <- f.swim.data %>%
                  mutate(Pool25m = ifelse(PoolConfiguration == "25m", 1, 0))

```

Fitting the Model with Jim's Code
```{r}
# First we need to add/create some of the variables to the df

# Season 
  # First we need to find the first year of performance for each athlete
f.swim.data <- f.swim.data %>%
  mutate(Year = year(ResultDate))
  # Then we can create a season variable in the df (season = 1 in the year of their first performance) 
f.swim.data <- f.swim.data %>%
  group_by(PersonId) %>%
  arrange(ResultDate) %>%
  mutate(
    Season = dense_rank(year(ResultDate))
  ) %>%
  ungroup()

# Times = The date of the performances within each season scaled to be between 0 and 1
  # First we need the day of the performance
f.swim.data <- f.swim.data %>%
  mutate(DayOfYear = yday(ResultDate))
  # Then scale within each season 
f.swim.data <- f.swim.data %>%
  group_by(Season) %>%
  mutate(
    Times = (DayOfYear - min(DayOfYear)) /
            (max(DayOfYear) - min(DayOfYear))
  ) %>%
  ungroup()

# Order the data properly 
  # (by athlete, then by season for each athlete, then by date within each season)
  # First create a numeric identifier for each athlete
f.swim.data <- f.swim.data %>%
  mutate(athlete = as.numeric(factor(PersonId)))

f.swim.data <- f.swim.data %>%
  arrange(athlete, Season, ResultDate)

# Create all of our variables needed for the fit_model function

# ages = Ages of all athletes
ages.f  <- split(f.swim.data$Age, f.swim.data$athlete)

# times = The date of the performances within each season scaled to be between 0 and 1
times.f <- split(f.swim.data$Times, f.swim.data$athlete)

# regressors_a = covariates for each performance
  # First extract the pool indicators
pool.f <- split(f.swim.data$Pool25m, f.swim.data$athlete)
  # Then add the age polynomial
mean_age <- mean(unlist(ages.f))
degree <- 4
  # Holder for the covariates
regressors_a.f <- rep(list(), length(ages.f))
for (i in seq_along(ages.f)) {
  # create matrix of 1s with degree+2 columns 
    # col 1: intercept (all 1s)
  regressors_a.f[[i]] <- matrix(1, nrow = length(ages.f[[i]]), ncol = degree + 2)
  # cols 2 to degree+1: polynomial terms (a - mean_age)^d
  for (j in 1:degree) {   
    regressors_a.f[[i]][, j + 1] <- (ages.f[[i]] - mean_age)^j
  }
  # col degree+2: pool indicator
  regressors_a.f[[i]][, degree + 2] <- pool.f[[i]]
}

# target = Performances of all athletes
target.f <- split(f.swim.data$TimeSeconds, f.swim.data$athlete)

# season = Season of each performance
season.f <- split(f.swim.data$Season, f.swim.data$athlete)

# direction = direction of improvement
  # Lower times = Better so performance improves by going down
direction <- "negative"

# MCMC Settings (tune these better later)
  # burnin = number of burn-in iterations
burnin <- 5000
  # numbofits = number of collected samples
numbofits <- 2000
  # every = thinning and so the sampler is run for burnin + numbofits * every iterations in total
every <- 10


# Split everything by athlete
ages.f <- split(f.swim.data$Age, f.swim.data$athlete)
times.f <- split(f.swim.data$Times, f.swim.data$athlete)
times.f <- lapply(times.f, function(x) {
  pmin(pmax(x, 1e-6), 1 - 1e-6)
})
season.f <- split(f.swim.data$Season, f.swim.data$athlete)
target.f <- split(f.swim.data$TimeSeconds, f.swim.data$athlete)


source("fit_model.R")
# fm.res <- fit_model(ages.f, times.f, regressors_a.f, target.f, season.f,
                    #direction, burnin, numbofits, every)
# Save the results
#saveRDS(fm.res, "fm_res.rds")
fm.res <- readRDS("fm_res.rds")

```

MCMC Convergence Check
```{r}
n_total <- length(fm.res$skew_alpha)  # Total number of iterations
n_each  <- n_total / 2                # Number of iterations per chain (since fit_model() runs 2 chains)
idx1  <- 1:n_each                     # Chain 1 indices
idx2  <- (n_each + 1):n_total         # Chain 2 indices

## Scalar parameters ##

# Put all of the vector parameters into one matrix 
scalar_params <- cbind(
  skew_alpha = fm.res$skew_alpha,
  nu1 = fm.res$nu1,
  nu2 = fm.res$nu2,
  nu_start = fm.res$nu_start,
  nu_inc = fm.res$nu_inc,
  sigmasq_inc = fm.res$sigmasq_inc,
  sigmasq_rw3 = fm.res$sigmasq_rw_3,
  sigmasq_start = fm.res$sigmasq_start,
  sigmasqmu = fm.res$sigmasqmu,
  sigmasqalpha = fm.res$sigmasqalpha
  #lambda0 = fm.res$a_sigmasqbeta_t_ind,   
  #lambda1 = fm.res$b_sigmasqbeta_t_ind,
  #tau0 = fm.res$a_sigmasqbeta_ind,
  #tau1 = fm.res$b_sigmasqbeta_ind
)
# scalar_params is [4000 x 10] -> we need to split it into the two chains
chain1_scalar <- mcmc(scalar_params[idx1, ])
chain2_scalar <- mcmc(scalar_params[idx2, ])
chains_scalar <- mcmc.list(chain1_scalar, chain2_scalar)

# PSRF and ESS for scalar parameters
psrf_scalar <- gelman.diag(chains_scalar, multivariate = FALSE)
ess_scalar <- effectiveSize(chains_scalar)

print(psrf_scalar)
print(ess_scalar)



## Matrix parameters ##
  # some parameters have different values for each athlete -> report worst case across rows (athletes)

# Helper function:
  # find the worst case across all athletes (high PSRF and low ESS) and report that
diagnose_matrix_param <- function(mat, idx1, idx2, param_name) {
  n_rows <- nrow(mat)          
  psrf_vals <- rep(NA, n_rows)    # Holder for the PSRF value of each athlete
  ess_vals <- rep(NA, n_rows)     # Holder for the ESS value of each athlete
  
  for (i in seq_len(n_rows)) {    
    ch <- mcmc.list(
      mcmc(mat[i, idx1]),         # Extract chain 1 for athlete i
      mcmc(mat[i, idx2])          # Extract chain 2 for athlete i 
    )                             # Then these are combined into an mcmc.list
    psrf_vals[i] <- gelman.diag(ch, multivariate = FALSE)$psrf[1, 1]    # Extract the point estimate of the PSRF value
    ess_vals[i] <- min(effectiveSize(ch))
  }
  
  cat(param_name, ": max PSRF =", round(max(psrf_vals), 3),
      " | min ESS =", round(min(ess_vals), 0), "\n")
  
  return(list(max_psrf = max(psrf_vals), min_ess = min(ess_vals)))  # return just the "worst case" 
}

# Run for each matrix parameter
diag_sigmasq <- diagnose_matrix_param(fm.res$sigmasq, idx1, idx2, "sigmasq")
diag_sigmasqbeta <- diagnose_matrix_param(fm.res$sigmasqbeta, idx1, idx2, "sigmasqbeta")
diag_sigmasqbeta_t <- diagnose_matrix_param(fm.res$sigmasqbeta_t, idx1, idx2, "sigmasqbeta_t")
diag_beta_b <- diagnose_matrix_param(fm.res$beta_b, idx1, idx2, "beta_b")
diag_beta_a <- diagnose_matrix_param(fm.res$beta_a, idx1, idx2, "beta_a")
diag_tau2 <- diagnose_matrix_param(fm.res$sigmasqbeta_ind, idx1, idx2, "tau2_i")
diag_lambda2 <- diagnose_matrix_param(fm.res$sigmasqbeta_t_ind, idx1, idx2, "lambda2_i")


## beta_bi ##

# This parameter is a list of 4000 and each element is [n_athletes x p_b] 
  # but we need it to be [n_athletes x p_b x 4000]
  # then we can extract one coefficient for one athlete across all iterations
n_athletes <- nrow(fm.res$sigmasq)
p_b <- nrow(fm.res$beta_b)

# Create a holder (of the correct structure) for the beta_bi
beta_bi_array <- array(NA, dim = c(n_athletes, p_b, n_total)) 

# Fill the holder
for (m in seq_len(n_total)) {                   # For each iteration m
  beta_bi_array[, , m] <- fm.res$beta_bi[[m]]   # Take that iteration's matrix and put it into the 3rd dim of the holder
}

# Similar to above, need to find worse case across all athlete-coefficient combos and report that 
  # Use similar method to the diagnose_matrix_param helper function 
psrf_bi <- rep(NA, n_athletes * p_b)            # Holder for the PSRF values
ess_bi <- rep(NA, n_athletes * p_b)             # Holder for the ESS values
k <- 1
for (i in seq_len(n_athletes)) {
  for (j in seq_len(p_b)) {
    ch <- mcmc.list(
      mcmc(beta_bi_array[i, j, idx1]),
      mcmc(beta_bi_array[i, j, idx2])
    )
    psrf_bi[k] <- gelman.diag(ch, multivariate = FALSE)$psrf[1, 1]
    ess_bi[k] <- min(effectiveSize(ch))
    k <- k + 1
  }
}
cat("beta_bi: max PSRF =", round(max(psrf_bi), 3),
    " | min ESS =", round(min(ess_bi), 0), "\n")


## Create summary table ##

diag_summary <- data.frame(
  Parameter = c(                                     # Vector of all the parameter names 
    "sigma^2_a", "sigma^2_m", "alpha", "zeta", 
    "beta_{n,v}", "beta^(i)_{n,v}", #"beta^(i,s)_{n,v}",
    "lambda^2_i", "c^2_{n,v}", "tau^2_i", "d^2_{n,v}", #"eta_{i,s}", 
    "sigma^2_i", "sigma^2_mu", "sigma^2_eta",
    "sigma^2_rw_3", "nu_1", "nu_2", "nu^mu", "nu^eta"),
    #"lambda_0", "lambda_1", "tau_0", "tau_1" ),
  Max_PSRF  = c(                                     # Vector of all the PSRF values
    psrf_scalar$psrf["sigmasqalpha",  1],
    psrf_scalar$psrf["sigmasqmu", 1],    
    psrf_scalar$psrf["skew_alpha",   1],
    diag_beta_a$max_psrf,
    diag_beta_b$max_psrf,
    max(psrf_bi),                                     # Just report the maximum PSRF of all the beta_bi 
    #NA, 
    diag_lambda2$max_psrf,
    diag_sigmasqbeta_t$max_psrf,
    diag_tau2$max_psrf,
    diag_sigmasqbeta$max_psrf,
    #NA,
    diag_sigmasq$max_psrf,
    psrf_scalar$psrf["sigmasq_start", 1],
    psrf_scalar$psrf["sigmasq_inc", 1],
    psrf_scalar$psrf["sigmasq_rw3", 1],
    psrf_scalar$psrf["nu1", 1],
    psrf_scalar$psrf["nu2", 1], 
    psrf_scalar$psrf["nu_start", 1],
    psrf_scalar$psrf["nu_inc", 1]),
    #psrf_scalar$psrf["lambda0", 1],
    #psrf_scalar$psrf["lambda1", 1],
    #psrf_scalar$psrf["tau0", 1],
    #psrf_scalar$psrf["tau1", 1]),
  Min_ESS   = c(                                       # Vector of all the ESS values
    ess_scalar["sigmasqalpha"],    
    ess_scalar["sigmasqmu"],    
    ess_scalar["skew_alpha"],
    diag_beta_a$min_ess,
    diag_beta_b$min_ess,
    min(ess_bi),                                       # Just report the minimum ESS of all the beta_bi
    #NA,
    diag_lambda2$min_ess,
    diag_sigmasqbeta_t$min_ess,
    diag_tau2$min_ess,
    diag_sigmasqbeta$min_ess,
    #NA,
    diag_sigmasq$min_ess,
    ess_scalar["sigmasq_start"],
    ess_scalar["sigmasq_inc"],
    ess_scalar["sigmasq_rw3"],
    ess_scalar["nu1"],
    ess_scalar["nu2"],
    ess_scalar["nu_start"],
    ess_scalar["nu_inc"])
    #ess_scalar["lambda0"],
    #ess_scalar["lambda1"],
    #ess_scalar["tau0"],
    #ess_scalar["tau1"])
)

print(diag_summary)



# Plot the first 500 iterations of chain 1 only
# to see if the chain has actually stabilised by then
par(mfrow = c(3, 2))
plot(fm.res$skew_alpha[1:500],  type = "l", main = expression(alpha))
plot(fm.res$nu1[1:500],         type = "l", main = expression(nu[1]))
plot(fm.res$nu2[1:500],         type = "l", main = expression(nu[2]))
plot(fm.res$sigmasq_inc[1:500], type = "l", main = expression(sigma[eta]^2))
plot(fm.res$nu_start[1:500],    type = "l", main = expression(nu^mu))
plot(fm.res$sigmasq[1, 1:500],  type = "l", main = "sigma^2 athlete 1")
par(mfrow = c(1, 1))

```

Single Athlete Prediction - LAYER 1
```{r}
set.seed(111)

#### LAYER 1 PREDICTION: g(a) only ####

athlete_idx <- 1
n_samples <- ncol(fm.res$beta_a)   # Number of MCMC samples 
mean_age <- mean(unlist(ages.f))   # grand mean age used in fitting
degree <- 4                        # degree of age polynomial

# Identify held out (last) season 
n_seasons <- length(unique(season.f[[athlete_idx]]))
last_season_idx <- which(season.f[[athlete_idx]] == n_seasons)

cat("Athlete", athlete_idx, "has", n_seasons, "seasons\n")
cat("Performances in last season:", length(last_season_idx), "\n")

# Actual (observed) values in last season
actual_times_last <- target.f[[athlete_idx]][last_season_idx]
actual_season_best <- min(actual_times_last)
ages_last <- ages.f[[athlete_idx]][last_season_idx]

cat("Actual times in last season:\n")
print(actual_times_last)
cat("Actual season best:", actual_season_best, "seconds\n")

# Prediction loop
  # For each MCMC sample, predict performance at each race in last season
    # then take minimum as predicted season best

  # Holder for the predicted values
pred_l1_all <- matrix(NA, nrow = n_samples, ncol = length(last_season_idx))

for (m in seq_len(n_samples)) {
  
  # Extract parameters for this MCMC sample
  beta_a_m <- fm.res$beta_a[, m]   # matrix of coefficients for 
                                      # [intercept, age^1, age^2, age^3, age^4, pool] for each sample m
  
  for (k in seq_along(last_season_idx)) {
    age_k <- ages_last[k]      # raw age of performance k
    age_c <- age_k - mean_age  # centered age of performance k
    des_mat <- c(1, age_c, age_c^2, age_c^3, age_c^4)   #design matrix
    pred_l1_all[m, k] <- as.numeric(des_mat %*% beta_a_m[1:5])  # predictions = design matrix * coefficients
  }
}

# Predicted season best = minimum across races in last season
pred_season_best_l1 <- apply(pred_l1_all, 1, min)

# Summary
cat("\n WOMEN'S ATHLETE", athlete_idx, "LAYER 1 RESULTS: \n")
cat("Actual season best:", round(actual_season_best, 3), "seconds\n")
cat("Predicted season best (median): ", round(median(pred_season_best_l1), 3), "seconds\n")
cat("95% credible interval: [",
    round(quantile(pred_season_best_l1, 0.025), 3), ",",
    round(quantile(pred_season_best_l1, 0.975), 3), "]\n")
cat("Actual inside 95% CI:",
    actual_season_best >= quantile(pred_season_best_l1, 0.025) &
    actual_season_best <= quantile(pred_season_best_l1, 0.975), "\n")
cat("Absolute error:",
    round(abs(median(pred_season_best_l1) - actual_season_best), 3), "seconds\n")

# Plot
hist(pred_season_best_l1,
     breaks = 50,
     main = "(W) Layer 1: Predicted Season Best (g(a) Only)",
     xlab = "Predicted Time (Seconds)",
     col = "steelblue", border = "white",
     xlim = range(c(pred_season_best_l1, actual_season_best)))
abline(v = actual_season_best,
       col = "red", lwd = 2, lty = 1)
abline(v = median(pred_season_best_l1),
       col = "black", lwd = 2, lty = 2)
abline(v = quantile(pred_season_best_l1, 0.025),
       col = "darkgrey", lwd = 1, lty = 3)
abline(v = quantile(pred_season_best_l1, 0.975),
       col = "darkgrey", lwd = 1, lty = 3)
legend("topleft",
       legend = c("Actual Season Best", "Predicted Median", "95% CI Bounds"),
       col = c("red", "black", "darkgrey"),
       lty = c(1, 2, 3), lwd = 2, bty = "n")
```

Single Athlete Prediction - LAYER 2
```{r}
#### LAYER 2 PREDICTION: g(a) + individual RW trend ####

# Create a function that extends F forward using a random walk
extend_F <- function(F_curr, n_future, sigmasq_inc, sigmasq_rw_3) {
  # F_curr = Season boundary values for athlete i from one MCMC sample (vector of length S+1)
  # n_future = Number of seasons we are predicting ahead
  # sigmasq_inc = Variance of the first increment of F 
  # sigmasq_rw_3 = Variance of the second differences of F 
  F_ext <- c(F_curr, rep(NA, n_future))    # Holder for the extended values
  S <- length(F_curr)
  
  for (s in (S + 1):(S + n_future)) { 
    if (s == 2) {
      # first increment variance 
        # only used when athlete has 1 observed season and we are predicting the 2nd season
      F_ext[s] <- F_ext[s - 1] + rnorm(1, mean = 0, sd = sqrt(sigmasq_inc))   
    } else {
      # second-difference variance: 
       # F[s] - 2*F[s-1] + F[s-2] ~ N(0, sigmasq_rw_3) => F[s] ~ N(2*F[s-1] - F[s-2], sigmasq_rw_3)
      F_ext[s] <- 2 * F_ext[s - 1] - F_ext[s - 2] + rnorm(1, mean = 0, sd = sqrt(sigmasq_rw_3))
    }
  }
  return(F_ext)
}


# Setup (same as Layer 1)
athlete_idx <- 1
n_samples <- ncol(fm.res$beta_a)
mean_age <- mean(unlist(ages.f))
degree <- 4                        

# Last season info (same as Layer 1)
n_seasons <- length(unique(season.f[[athlete_idx]]))
last_season_idx <- which(season.f[[athlete_idx]] == n_seasons)
actual_times_last <- target.f[[athlete_idx]][last_season_idx]
actual_season_best <- min(actual_times_last)
#pool_last <- regressors_a.f[[athlete_idx]][last_season_idx, 2]
ages_last <- ages.f[[athlete_idx]][last_season_idx]
z_last <- times.f[[athlete_idx]][last_season_idx]  # within-season positions of 
                                                     # each performance in the last season

# Prediction loop
  # Holder for the predicted values 
pred_l2_all <- matrix(NA, nrow = n_samples, ncol = length(last_season_idx))

for (m in seq_len(n_samples)) {
  
  # Extract parameters for this MCMC sample
  beta_a_m <- fm.res$beta_a[, m]
  sigmasq_inc_m <- fm.res$sigmasq_inc[m]    # variance of the first increment
  sigmasq_rw3_m <- fm.res$sigmasq_rw_3[m]   # variance of the second difference of F
  
  # Extract F values up to and including start of last season
    # In case of Athlete 1: the "observed" boundaries are rows 1 to 9 (seasons 1-8)
    # i.e. eta_{1,1} to eta_{1,9} (end of season 8)
  F_observed <- fm.res$F[[athlete_idx]][1:n_seasons, m]
  
  # Extend F forward by 1 step 
    # In case of Athlete 1: extend F forward from season 8 to get season 9 boundaries (to get eta_{1,10})
  F_extended <- extend_F(F_observed, n_future = 1, sigmasq_inc = sigmasq_inc_m,
                         sigmasq_rw_3 = sigmasq_rw3_m)
  
  # "Next" season boundary values (in case of Athlete 1: Season 9)
  eta_start <- F_extended[n_seasons]       # in case of Athlete 1: eta_{1,9} = start of season 9
  eta_end <- F_extended[n_seasons + 1]     # and eta_{1,10} = end of season 9
  
  for (k in seq_along(last_season_idx)) {
    
    # Layer 1 component: g(a)
    age_k <- ages_last[k]
    age_c <- age_k - mean_age
    des_mat <- c(1, age_c, age_c^2, age_c^3, age_c^4)
    layer1_k <- as.numeric(des_mat %*% beta_a_m[1:5])           # layer 1 prediction 
    
    # Layer 2 component: piecewise linear interpolation of F
    z_k <- z_last[k]   # within-season position of this performance (between 0 and 1)
    fi_z_k <- (1 - z_k) * eta_start + z_k * eta_end  # "trend excess performance trajectory"
    
    # Combined prediction = [g(a)] + [individual RW trend]
    pred_l2_all[m, k] <- layer1_k + fi_z_k
  }
}

# Predicted season best = minimum across performances in last season
pred_season_best_l2 <- apply(pred_l2_all, 1, min)

# Summary
cat("\n WOMEN'S ATHLETE", athlete_idx, "LAYER 2 RESULTS \n")
cat("Actual season best:", round(actual_season_best, 3), "seconds\n")
cat("Predicted season best (median):", round(median(pred_season_best_l2), 3), "seconds\n")
cat("95% credible interval: [",
    round(quantile(pred_season_best_l2, 0.025), 3), ",",
    round(quantile(pred_season_best_l2, 0.975), 3), "]\n")
cat("Actual inside 95% CI:",
    actual_season_best >= quantile(pred_season_best_l2, 0.025) &
    actual_season_best <= quantile(pred_season_best_l2, 0.975), "\n")
cat("Absolute error:", round(abs(median(pred_season_best_l2) - actual_season_best), 3), "seconds\n")

# Compare Layer 1 vs Layer 2
cat("\n COMPARISON OF LAYER 1 VS LAYER 2 \n")
cat("Layer 1 absolute error:", 
    round(abs(median(pred_season_best_l1) - actual_season_best), 3), "seconds\n")
cat("Layer 2 absolute error:", 
    round(abs(median(pred_season_best_l2) - actual_season_best), 3), "seconds\n")
cat("Layer 1 CI width:", 
    round(diff(quantile(pred_season_best_l1, c(0.025, 0.975))), 3), "seconds\n")
cat("Layer 2 CI width:", 
    round(diff(quantile(pred_season_best_l2, c(0.025, 0.975))), 3), "seconds\n")

# Plot comparing Layer 1 and Layer 2
par(mfrow = c(1, 2))

hist(pred_season_best_l1,
     breaks = 50, col = "plum", border = "white",
     main = "(W) Layer 1: g(a) Only",
     xlab = "Predicted Time (Seconds)", xlim = range(c(pred_season_best_l1, actual_season_best)))
abline(v = actual_season_best, col = "red", lwd = 2)
abline(v = median(pred_season_best_l1), col = "black", lwd = 2, lty = 2)
abline(v = quantile(pred_season_best_l1, 0.025), col = "darkgrey", lwd = 1, lty = 3)
abline(v = quantile(pred_season_best_l1, 0.975), col = "darkgrey", lwd = 1, lty = 3)

hist(pred_season_best_l2,
     breaks = 50, col = "seagreen", border = "white",
     main = "(W) Layer 2: + Individual RW Trend",
     xlab = "Predicted Time (Seconds)", xlim = range(c(pred_season_best_l2, actual_season_best)))
abline(v = actual_season_best, col = "red", lwd = 2)
abline(v = median(pred_season_best_l2), col = "black", lwd = 2, lty = 2)
abline(v = quantile(pred_season_best_l2, 0.025), col = "darkgrey", lwd = 1, lty = 3)
abline(v = quantile(pred_season_best_l2, 0.975), col = "darkgrey", lwd = 1, lty = 3)
legend("topleft",
       legend = c("Actual Season Best", "Predicted Median", "95% CI Bounds"),
       col = c("red", "black", "darkgrey"),
       lty = c(1, 2, 3), lwd = 2, bty = "n")

```

Single Athlete Prediction - LAYER 3
```{r}
#### LAYER 3 PREDICTION ####
## full model -> g(a) + individual RW trend + within season shape

# Setup (same as Layers 1 and 2)
athlete_idx <- 1
n_samples <- ncol(fm.res$beta_a)
mean_age <- mean(unlist(ages.f))   
degree <- 4                       
p_b <- nrow(fm.res$beta_b)             # Number of RBP basis functions

# Last season info (same as Layer 2)
n_seasons <- length(unique(season.f[[athlete_idx]]))
last_season_idx <- which(season.f[[athlete_idx]] == n_seasons)
actual_times_last <- target.f[[athlete_idx]][last_season_idx]
actual_season_best <- min(actual_times_last)
ages_last <- ages.f[[athlete_idx]][last_season_idx]
z_last <- times.f[[athlete_idx]][last_season_idx]

# Construct the RBP for the within-season time points
  # This is the same TruncBernstein function used in fit_model()
b_last <- TruncBernstein(z_last, degree = 4)

# Prediction loop
  # Holder for the predicted values 
pred_l3_all <- matrix(NA, nrow = n_samples, ncol = length(last_season_idx))

for (m in seq_len(n_samples)) {
  
  # Extract parameters for this MCMC sample
  beta_a_m <- fm.res$beta_a[, m]
  sigmasq_inc_m <- fm.res$sigmasq_inc[m]
  sigmasq_rw3_m <- fm.res$sigmasq_rw_3[m]
  beta_bi_m <- fm.res$beta_bi[[m]][athlete_idx, ]                  # athlete-level RBP coefficients 
  sigmasqbeta_t_m <- fm.res$sigmasqbeta_t[, m]                     # hyperparameter c^2
  sigmasqbeta_t_ind_m <- fm.res$sigmasqbeta_t_ind[athlete_idx, m]  # lambda^2_i
  
  # Extend F forward (same as Layer 2)
  F_observed <- fm.res$F[[athlete_idx]][1:n_seasons, m]
  F_extended <- extend_F(F_observed, n_future = 1, sigmasq_inc = sigmasq_inc_m,
                         sigmasq_rw_3 = sigmasq_rw3_m)
  eta_start <- F_extended[n_seasons]
  eta_end <- F_extended[n_seasons + 1]
  
  # Draw new within-season RBP coefficients for the predicted season
    # beta_bit_new ~ N(beta_bi_i, lambda^2_i * c^2_{n,v})
  beta_bit_new <- rnorm(p_b, mean = beta_bi_m, sd = sqrt(sigmasqbeta_t_ind_m * sigmasqbeta_t_m))
  
  # Evaluate within-season shape at each performance time point
    # h*_{i,s}(z) = b_last * beta_bit_new
  h_star <- as.numeric(b_last %*% beta_bit_new)   # length = n_performances = 5
  
  for (k in seq_along(last_season_idx)) {
    
    # Layer 1 component: g(a) only
    age_k <- ages_last[k]
    age_c <- age_k - mean_age
    des_mat <- c(1, age_c, age_c^2, age_c^3, age_c^4)
    layer1_k <- as.numeric(des_mat %*% beta_a_m[1:5])
    
    # Layer 2 component: piecewise linear interpolation of F
    z_k <- z_last[k]
    fi_z_k <- (1 - z_k) * eta_start + z_k * eta_end
    
    # Layer 3 component: within-season shape
    h_star_k <- h_star[k]
    
    # Combined prediction = [g(a)] + [individual RW trend] + [within-season shape]
    pred_l3_all[m, k] <- layer1_k + fi_z_k + h_star_k
  }
}

# Predicted season best = minimum across performances in last season
pred_season_best_l3 <- apply(pred_l3_all, 1, min)


# Summary
cat("\n WOMEN'S ATHLETE", athlete_idx, "LAYER 3 RESULTS \n")
cat("Actual season best:", round(actual_season_best, 3), "seconds\n")
cat("Predicted season best (median):", round(median(pred_season_best_l3), 3), "seconds\n")
cat("95% credible interval: [",
    round(quantile(pred_season_best_l3, 0.025), 3), ",",
    round(quantile(pred_season_best_l3, 0.975), 3), "]\n")
cat("Actual inside 95% CI:",
    actual_season_best >= quantile(pred_season_best_l3, 0.025) &
    actual_season_best <= quantile(pred_season_best_l3, 0.975), "\n")
cat("Absolute error:", round(abs(median(pred_season_best_l3) - actual_season_best), 3), "seconds\n")

# Full comparison across all three layers
cat("\nFULL COMPARISON \n")
cat(sprintf("%-10s %-10s %-10s %-10s %-10s\n", "Layer", "Median", "Error", "CI Width", "In CI?"))
cat(sprintf("%-10s %-10f %-10f %-10f %-10s\n",
            "Layer 1",
            median(pred_season_best_l1),
            abs(median(pred_season_best_l1) - actual_season_best),
            diff(quantile(pred_season_best_l1, c(0.025, 0.975))),
            actual_season_best >= quantile(pred_season_best_l1, 0.025) &
            actual_season_best <= quantile(pred_season_best_l1, 0.975)))
cat(sprintf("%-10s %-10f %-10f %-10f %-10s\n",
            "Layer 2",
            median(pred_season_best_l2),
            abs(median(pred_season_best_l2) - actual_season_best),
            diff(quantile(pred_season_best_l2, c(0.025, 0.975))),
            actual_season_best >= quantile(pred_season_best_l2, 0.025) &
            actual_season_best <= quantile(pred_season_best_l2, 0.975)))
cat(sprintf("%-10s %-10f %-10f %-10f %-10s\n",
            "Layer 3",
            median(pred_season_best_l3),
            abs(median(pred_season_best_l3) - actual_season_best),
            diff(quantile(pred_season_best_l3, c(0.025, 0.975))),
            actual_season_best >= quantile(pred_season_best_l3, 0.025) &
            actual_season_best <= quantile(pred_season_best_l3, 0.975)))

# Plot all three layers side by side
par(mfrow = c(1, 3))

hist(pred_season_best_l1,
     breaks = 50, col = "plum", border = "white",
     main = "(W) Layer 1:\ng(a)",
     xlab = "Predicted time (seconds)", xlim = range(c(pred_season_best_l1,actual_season_best)))
abline(v = actual_season_best, col = "red", lwd = 2)
abline(v = median(pred_season_best_l1), col = "black", lwd = 2, lty = 2)
abline(v = quantile(pred_season_best_l1, 0.025), col = "darkgrey", lwd = 1, lty = 3)
abline(v = quantile(pred_season_best_l1, 0.975), col = "darkgrey", lwd = 1, lty = 3)
legend("topleft",
       legend = c("Actual Season Best", "Predicted Median", "95% CI Bounds"),
       col = c("red", "black", "darkgrey"),
       lty = c(1, 2, 3), lwd = 2, bty = "n", cex=0.7)

hist(pred_season_best_l2,
     breaks = 50, col = "seagreen", border = "white",
     main = "(W) Layer 2:\n+ Individual RW Trend",
     xlab = "Predicted Time (Seconds)", xlim = range(c(pred_season_best_l2,actual_season_best)))
abline(v = actual_season_best, col = "red", lwd = 2)
abline(v = median(pred_season_best_l2), col = "black",  lwd = 2, lty = 2)
abline(v = quantile(pred_season_best_l2, 0.025), col = "darkgrey", lwd = 1, lty = 3)
abline(v = quantile(pred_season_best_l2, 0.975), col = "darkgrey", lwd = 1, lty = 3)

hist(pred_season_best_l3,
     breaks = 50, col = "steelblue", border = "white",
     main = "(W) Layer 3:\n+ Within-Season Shape",
     xlab = "Predicted Time (Seconds)", xlim = range(c(pred_season_best_l3,actual_season_best)))
abline(v = actual_season_best, col = "red", lwd = 2)
abline(v = median(pred_season_best_l3), col = "black",  lwd = 2, lty = 2)
abline(v = quantile(pred_season_best_l3, 0.025), col = "darkgrey", lwd = 1, lty = 3)
abline(v = quantile(pred_season_best_l3, 0.975), col = "darkgrey", lwd = 1, lty = 3)



### Using Layer 3 to Predict Timing of Peak Performance ### 

# Evaluate the Layer 3 prediction across a fine grid of z values
  # to find when within the season the predicted peak occurs

z_grid <- seq(0.01, 0.99, length.out = 200)    # fine grid of time points across season
                                                 # exclude boundaries where h*=0 by construction
b_grid <- TruncBernstein(z_grid, degree = 4)   # evaluate RBP basis at all grid points
ages_last <- ages.f[[athlete_idx]][last_season_idx]
future_age <- mean(ages_last) + 1    # athlete's approximate age in the future season
age_future_c <- future_age - mean_age  # centered future age

peak_z_samples <- rep(NA, n_samples)   # holder for predicted peak timing per MCMC sample

for (m in seq_len(n_samples)) {
  
  beta_a_m <- fm.res$beta_a[, m]
  sigmasq_inc_m <- fm.res$sigmasq_inc[m]
  sigmasq_rw3_m <- fm.res$sigmasq_rw_3[m]
  beta_bi_m <- fm.res$beta_bi[[m]][athlete_idx, ]
  sigmasqbeta_t_m <- fm.res$sigmasqbeta_t[, m]
  sigmasqbeta_t_ind_m <- fm.res$sigmasqbeta_t_ind[athlete_idx, m]
  
  # Extend F (same as before)
  F_observed <- fm.res$F[[athlete_idx]][1:n_seasons, m]
  F_extended <- extend_F(F_observed, n_future = 1, sigmasq_inc = sigmasq_inc_m,
                         sigmasq_rw_3 = sigmasq_rw3_m)
  eta_start <- F_extended[n_seasons]
  eta_end <- F_extended[n_seasons + 1]
  
  # Draw within-season coefficients
  beta_bit_new <- rnorm(p_b, mean = beta_bi_m, sd = sqrt(sigmasqbeta_t_ind_m * sigmasqbeta_t_m))
  
  # Evaluate each component of the full prediction across the fine grid
  layer1_grid <- as.numeric(c(1, age_future_c, age_future_c^2, age_future_c^3, 
                               age_future_c^4) %*% beta_a_m[1:5])
  fi_grid <- (1 - z_grid) * eta_start + z_grid * eta_end
  h_star_grid <- as.numeric(b_grid %*% beta_bit_new)
  
  # Combine into full predicted performance curve for this MCMC sample
  pred_perform_curve <- layer1_grid + fi_grid + h_star_grid   
  
  # Peak = minimum predicted time
  peak_z_samples[m] <- z_grid[which.min(pred_perform_curve)]
}

# Convert z to approximate month (z=0 = Jan, z=1 = Dec)
month_breaks <- seq(0, 1, length.out = 13)
month_labels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")
peak_month_samples <- cut(peak_z_samples, breaks = month_breaks, labels = month_labels)

# Summary of peak timing
cat("\nWOMEN'S ATHLETE", athlete_idx, "PREDICTED PEAK TIMING\n")
cat("Predicted peak z (median):", round(median(peak_z_samples), 3), "\n")
cat("95% CI for peak z:[",
    round(quantile(peak_z_samples, 0.025), 3), ",",
    round(quantile(peak_z_samples, 0.975), 3), "]\n")
cat("\nPredicted peak month distribution:\n")
print(table(peak_month_samples))

# Plot peak month distribution
barplot(table(peak_month_samples) / n_samples, col = "steelblue",
        main = "(W) Predicted Peak Performance Month - Athlete 1",
        xlab = "Month", ylab   = "Probability", las = 2)

```

GENERIC GRAPHS (ie: can be used on any athlete)
```{r}
## GENERIC VERSION ##
#### Multi-Season Horizon with Intermediate Predictions ####
  # ie: one prediction per month
  # for 1, 2, and 3 seasons ahead
  # these leave out the 25m pool performances
## FOR LAYER 3 MODEL ## 

# Setup variables
athlete_idx <- 1
n_samples <- ncol(fm.res$beta_a)
mean_age <- mean(unlist(ages.f))
degree <- 4
n_seasons <- length(unique(season.f[[athlete_idx]]))
target_season <- n_seasons

a1 <- n_seasons - 2
a2 <- n_seasons - 1
a3 <- n_seasons

athlete_rows_x <- f.swim.data$athlete == athlete_idx
athlete_id_x <- unique(f.swim.data$PersonId[athlete_rows_x])

# Compute true exact ages for all athletes (before we only had integer age)
f.swim.full.data$DateOfBirth <- as.Date(f.swim.full.data$DateOfBirth)
f.swim.full.data$ResultDate <- as.Date(f.swim.full.data$ResultDate)
f.swim.full.data$exact_age_true <- as.numeric(
  f.swim.full.data$ResultDate - f.swim.full.data$DateOfBirth) / 365.25

  # exact ages for Athlete x
exact_age_x <- f.swim.full.data$exact_age_true[
  match(
    paste(f.swim.data$ResultDate[athlete_rows_x],
          f.swim.data$PersonId[athlete_rows_x]),
    paste(f.swim.full.data$ResultDate,
          f.swim.full.data$PersonId.x)
  )
]
  # date of birth for Athlete x
dob_x <- as.Date(unique(f.swim.full.data$DateOfBirth[
  f.swim.full.data$PersonId.x == athlete_id_x]))

cat("Exact age range:", round(min(exact_age_x), 2),
    "to", round(max(exact_age_x), 2), "\n")
cat("Date of birth:", format(dob_x), "\n")
cat("Any missing:", any(is.na(exact_age_x)), "\n")

# Check that there are no gap years
  # (we will also need season_years_x when running the model)
season_years_x <- sapply(1:length(unique(season.f[[athlete_idx]])), function(s) {
  season_rows <- which(season.f[[athlete_idx]] == s)
  as.numeric(names(sort(table(
    f.swim.data$Year[f.swim.data$athlete == athlete_idx][season_rows]
  ), decreasing = TRUE))[1])
})
cat("Season years:", season_years_x, "\n")

dob <- dob_x
exact_age <- exact_age_x
season_years <- season_years_x

# Recompute season age ranges for current athlete
season_start_ages <- sapply(season_years, function(yr) {
  as.numeric(as.Date(paste0(yr, "-01-01")) - dob) / 365.25
})
season_end_ages <- sapply(season_years, function(yr) {
  as.numeric(as.Date(paste0(yr + 1, "-01-01")) - dob) / 365.25
})
season_gaps <- diff(season_years) > 1
years_observed <- season_years

# Recompute fitted curves for Athlete x
z_fine <- seq(0, 1, length.out = 102)
b_fine <- TruncBernstein(z_fine, degree = 4)
z_monthly <- seq(1/24, 23/24, length.out = 12)
p_b <- nrow(fm.res$beta_b) 

fitted_fine <- vector("list", n_seasons)
for (s in 1:n_seasons) {
  fitted_fine[[s]] <- matrix(NA, nrow = n_samples, ncol = length(z_fine))
}

# Compute fitted curves for all seasons
for (m in seq_len(n_samples)) {
  beta_a_m <- fm.res$beta_a[, m]
  
  for (s in 1:n_seasons) {
    ages_z <- season_start_ages[s] + z_fine * (season_end_ages[s] - season_start_ages[s])
    beta_bit_s <- fm.res$beta_bit[[athlete_idx]][[m]][s, ]
    eta_s_start <- fm.res$F[[athlete_idx]][s, m]
    eta_s_end <- fm.res$F[[athlete_idx]][s + 1, m]
    
    for (j in seq_along(z_fine)) {
      age_c <- ages_z[j] - mean_age
      des_mat <- c(1, age_c, age_c^2, age_c^3, age_c^4)
      g_a <- as.numeric(des_mat %*% beta_a_m[1:5])
      fi_z <- (1 - z_fine[j]) * eta_s_start + z_fine[j] * eta_s_end
      h_star <- as.numeric(b_fine[j, ] %*% beta_bit_s)
      fitted_fine[[s]][m, j] <- g_a + fi_z + h_star
    }
  }
}

# Summary of fitted curves
fitted_fine_median <- lapply(1:n_seasons, function(s)
  apply(fitted_fine[[s]], 2, median))
fitted_fine_lower <- lapply(1:n_seasons, function(s)
  apply(fitted_fine[[s]], 2, quantile, 0.025))
fitted_fine_upper <- lapply(1:n_seasons, function(s)
  apply(fitted_fine[[s]], 2, quantile, 0.975))

# Helper function: computes predicted curve at monthly points
  # for a given horizon, starting from a given season
predict_season_curve <- function(fm.res, athlete_idx, train_up_to, 
                                  predict_season, n_samples, p_b,
                                  mean_age, z_monthly, ages_monthly_pred,
                                  sigmasq_inc_vec, sigmasq_rw3_vec) {
  
  # holder for predictions
  pred_matrix <- matrix(NA, nrow = n_samples, ncol = 12)
  
  # prediction loop (same as before)
  for (m in seq_len(n_samples)) {
    
    beta_a_m <- fm.res$beta_a[, m]
    sigmasq_inc_m <- sigmasq_inc_vec[m]
    sigmasq_rw3_m <- sigmasq_rw3_vec[m]
    beta_bi_m <- fm.res$beta_bi[[m]][athlete_idx, ]
    sigmasqbeta_t_m <- fm.res$sigmasqbeta_t[, m]
    sigmasqbeta_t_ind_m <- fm.res$sigmasqbeta_t_ind[athlete_idx, m]
    
    # extend F from end of training seasons to target season
    n_future <- predict_season - train_up_to
    F_obs <- fm.res$F[[athlete_idx]][1:train_up_to, m]
    F_ext <- extend_F(F_obs, n_future = n_future, sigmasq_inc  = sigmasq_inc_m,
                           sigmasq_rw_3 = sigmasq_rw3_m)
    eta_start <- F_ext[predict_season - 1]
    eta_end <- F_ext[predict_season] 
    
    # draw within-season coefficients for target season
    beta_bit_new <- rnorm(p_b, mean = beta_bi_m, sd = sqrt(sigmasqbeta_t_ind_m * sigmasqbeta_t_m))
    
    # predictions for each month
    for (j in 1:12) {
      age_c <- ages_monthly_pred[j] - mean_age
      des_mat <- c(1, age_c, age_c^2, age_c^3, age_c^4)
      g_a <- as.numeric(des_mat %*% beta_a_m[1:5])
      fi_z <- (1 - z_monthly[j]) * eta_start + z_monthly[j] * eta_end
      b_z <- TruncBernstein(z_monthly[j], degree = 4)
      h_star <- as.numeric(b_z %*% beta_bit_new)
      pred_matrix[m, j] <- g_a + fi_z + h_star
    }
  }
  
  # return median + upper/lower bounds of 95% CI
  list(median = apply(pred_matrix, 2, median),
    lower = apply(pred_matrix, 2, quantile, 0.025),
    upper = apply(pred_matrix, 2, quantile, 0.975))
}


# Helper function: computes ages_monthly for each predicted season
get_ages_monthly <- function(season_yr, dob, z_monthly) {
  age_jan1  <- as.numeric(as.Date(paste0(season_yr, "-01-01")) - dob) / 365.25
  age_dec31 <- as.numeric(as.Date(paste0(season_yr, "-12-31")) - dob) / 365.25
  age_jan1 + z_monthly * (age_dec31 - age_jan1)
}

# Compute ages_monthly for 1, 2, 3 seasons ahead
ages_monthly_s1 <- get_ages_monthly(season_years[a1], dob, z_monthly)
ages_monthly_s2 <- get_ages_monthly(season_years[a2], dob, z_monthly)
ages_monthly_s3 <- get_ages_monthly(season_years[a3], dob, z_monthly)


# Use helper function to compute all predicted curves 

# Horizon 1: train on all but one season, predict the last one
pred_h1_s1 <- predict_season_curve(
  fm.res, athlete_idx, train_up_to = (a3 - 1), predict_season = a3,
  n_samples, p_b, mean_age, z_monthly, ages_monthly_s3,
  fm.res$sigmasq_inc, fm.res$sigmasq_rw_3)

# Horizon 2: train on all by two seasons, predict the last two
pred_h2_s1 <- predict_season_curve(
  fm.res, athlete_idx, train_up_to = (a3 - 2), predict_season = (a3 - 1),
  n_samples, p_b, mean_age, z_monthly, ages_monthly_s2,
  fm.res$sigmasq_inc, fm.res$sigmasq_rw_3)

pred_h2_s2 <- predict_season_curve(
  fm.res, athlete_idx, train_up_to = (a3 - 2), predict_season = a3,
  n_samples, p_b, mean_age, z_monthly, ages_monthly_s3,
  fm.res$sigmasq_inc, fm.res$sigmasq_rw_3)

# Horizon 3: train on all but three seasons, predict the last three
pred_h3_s1 <- predict_season_curve(
  fm.res, athlete_idx, train_up_to = (a3 - 3), predict_season = (a3 - 2),
  n_samples, p_b, mean_age, z_monthly, ages_monthly_s1,
  fm.res$sigmasq_inc, fm.res$sigmasq_rw_3)

pred_h3_s2 <- predict_season_curve(
  fm.res, athlete_idx, train_up_to = (a3 - 3), predict_season = (a3 - 1),
  n_samples, p_b, mean_age, z_monthly, ages_monthly_s2,
  fm.res$sigmasq_inc, fm.res$sigmasq_rw_3)

pred_h3_s3 <- predict_season_curve(
  fm.res, athlete_idx, train_up_to = (a3 - 3), predict_season = a3,
  n_samples, p_b, mean_age, z_monthly, ages_monthly_s3,
  fm.res$sigmasq_inc, fm.res$sigmasq_rw_3)


# Plots

# Helper function: draw one season band 
draw_season_band <- function(season_yr, dob, y_min, y_max, col) {
  age_jan1  <- as.numeric(as.Date(paste0(season_yr, "-01-01")) - dob) / 365.25
  age_dec31 <- as.numeric(as.Date(paste0(season_yr, "-12-31")) - dob) / 365.25
  rect(xleft = age_jan1, xright = age_dec31,
       ybottom = y_min, ytop = y_max,
       col = col, border = NA)
}

# Helper function: draw prediction CI
draw_pred_ci <- function(ages_monthly, pred, col_ci) {
  polygon(c(ages_monthly, rev(ages_monthly)),
          c(pred$lower, rev(pred$upper)),
          col = col_ci, border = NA)
}

# Helper function: draw prediction median line
draw_pred_med <- function(ages_monthly, pred) {
  lines(ages_monthly, pred$median, col = "black", lwd = 2, lty = 2)
}


# Shared plot settings
all_preds <- c(pred_h1_s1$lower, pred_h1_s1$upper,
               pred_h2_s1$lower, pred_h2_s1$upper,
               pred_h2_s2$lower, pred_h2_s2$upper,
               pred_h3_s1$lower, pred_h3_s1$upper,
               pred_h3_s2$lower, pred_h3_s2$upper,
               pred_h3_s3$lower, pred_h3_s3$upper)

y_min_h <- min(c(target.f[[athlete_idx]], all_preds)) - 0.5
y_max_h <- max(c(target.f[[athlete_idx]], all_preds)) + 0.5
x_min_h <- season_start_ages[1] - 0.1
x_max_h <- as.numeric(as.Date(paste0(season_years[a3], "-12-31")) - dob) / 365.25 + 0.1

# Filter to take out the 25m pool performances (only want 50m pool ones)
pool_indicator <- regressors_a.f[[athlete_idx]][, degree + 2]
pool_50m_idx <- which(pool_indicator == 0)


#par(mfrow = c(1, 3))

# Horizon 1
plot(NA, NA,
     xlim = c(x_min_h, x_max_h), ylim = c(y_min_h, y_max_h),
     xlab = "Age (Years)", ylab = "Performance (Seconds)",
     main = "1 Season Ahead")

  # calendar year bands
for (y_idx in seq_along(years_observed)) {
  yr <- years_observed[y_idx]
  if (y_idx %% 2 == 1)
    draw_season_band(yr, dob, y_min_h, y_max_h, "grey90")
}

  # band for future season
draw_season_band(season_years[a3], dob, y_min_h, y_max_h, "lightblue1")

  # prediction CI for one season ahead
draw_pred_ci(ages_monthly_s3, pred_h1_s1, "lightblue3")

  # fitted CI and median line for all seasons
for (s in 1:a3) {
  ages_z_s <- season_start_ages[s] + z_fine * (season_end_ages[s] - season_start_ages[s])
  polygon(c(ages_z_s, rev(ages_z_s)),
          c(fitted_fine_lower[[s]], rev(fitted_fine_upper[[s]])),
          col = "grey70", border = NA)
  lines(ages_z_s, fitted_fine_median[[s]], col = "black", lwd = 1.5)
}

  # prediction median line for one season ahead
draw_pred_med(ages_monthly_s3, pred_h1_s1)

  # observed performances
obs_train <- which(season.f[[athlete_idx]] <= (a3 - 1))
obs_test <- which(season.f[[athlete_idx]] == a3)
  # keep only the 50m pool performances
obs_train_50m <- intersect(obs_train, pool_50m_idx)
obs_test_50m <- intersect(obs_test,  pool_50m_idx)
points(exact_age[obs_train_50m], target.f[[athlete_idx]][obs_train_50m],
       pch = 19, cex = 0.6, col = "black")
points(exact_age[obs_test_50m], target.f[[athlete_idx]][obs_test_50m],
       pch = 1, cex = 0.6)

legend("bottomleft",
       legend = c("Training performances",
                  "Held-out performances",
                  "Fitted trajectory (median)",
                  "Fitted 95% CI",
                  "Predicted trajectory (median)",
                  "Predicted 95% CI"),
       col = c("black", "black", "black", "grey70", "black", "lightblue3"),
       lty = c(NA, NA, 1, NA, 2, NA), pch = c(19, 1, NA, 15, NA, 15),
       lwd = c(NA, NA, 1.5, NA, 2, NA), bty = "n", cex = 0.6)


# Horizon 2
plot(NA, NA,
     xlim = c(x_min_h, x_max_h), ylim = c(y_min_h, y_max_h),
     xlab = "Age (Years)", ylab = "Performance (Seconds)",
     main = "2 Seasons Ahead")

for (y_idx in seq_along(years_observed)) {
  yr <- years_observed[y_idx]
  if (y_idx %% 2 == 1)
    draw_season_band(yr, dob, y_min_h, y_max_h, "grey90")
}

  # prediction bands for 1 and 2 seasons ahead
draw_season_band(season_years[a2], dob, y_min_h, y_max_h, "lightblue1")
draw_season_band(season_years[a3], dob, y_min_h, y_max_h, "lightblue2")

  # prediction CI for seasons 8 and 9
draw_pred_ci(ages_monthly_s2, pred_h2_s1, "lightblue3")
draw_pred_ci(ages_monthly_s3, pred_h2_s2, "lightblue3")

  # fitted CI and median line for all seasons
for (s in 1:a3) {
  ages_z_s <- season_start_ages[s] + z_fine * (season_end_ages[s] - season_start_ages[s])
  polygon(c(ages_z_s, rev(ages_z_s)),
          c(fitted_fine_lower[[s]], rev(fitted_fine_upper[[s]])),
          col = "grey70", border = NA)
  lines(ages_z_s, fitted_fine_median[[s]], col = "black", lwd = 1.5)
}

  # prediction median lines for seasons 8 and 9
draw_pred_med(ages_monthly_s2, pred_h2_s1)
draw_pred_med(ages_monthly_s3, pred_h2_s2)

  # observed performances
obs_train <- which(season.f[[athlete_idx]] <= a1)
obs_test <- which(season.f[[athlete_idx]] %in% c(a2, a3))
obs_train_50m <- intersect(obs_train, pool_50m_idx)
obs_test_50m  <- intersect(obs_test,  pool_50m_idx)
points(exact_age[obs_train_50m], target.f[[athlete_idx]][obs_train_50m],
       pch = 19, cex = 0.6, col = "black")
points(exact_age[obs_test_50m], target.f[[athlete_idx]][obs_test_50m],
       pch = 1, cex = 0.6)

legend("bottomleft",
       legend = c("Training performances",
                  "Held-out performances",
                  "Fitted trajectory (median)",
                  "Fitted 95% CI",
                  "Predicted trajectory (median)",
                  "Predicted 95% CI"),
       col = c("black", "black", "black", "grey70", "black","lightblue3"),
       lty = c(NA, NA, 1, NA, 2, NA), pch = c(19, 1, NA, 15, NA, 15),
       lwd = c(NA, NA, 1.5, NA, 2, NA), bty = "n", cex = 0.6)


# Horizon 3
plot(NA, NA,
     xlim = c(x_min_h, x_max_h), ylim = c(y_min_h, y_max_h),
     xlab = "Age (Years)", ylab = "Performance (Seconds)",
     main = "3 Seasons Ahead")

for (y_idx in seq_along(years_observed)) {
  yr <- years_observed[y_idx]
  if (y_idx %% 2 == 1)
    draw_season_band(yr, dob, y_min_h, y_max_h, "grey90")
}

  # prediction bands for 1,2,3 seasons ahead
draw_season_band(season_years[a1], dob, y_min_h, y_max_h, "lightblue1")
draw_season_band(season_years[a2], dob, y_min_h, y_max_h, "lightblue2")
draw_season_band(season_years[a3], dob, y_min_h, y_max_h, "lightblue1")

  # prediction CI for 1,2,3 seasons ahead
draw_pred_ci(ages_monthly_s1, pred_h3_s1, "lightblue3")
draw_pred_ci(ages_monthly_s2, pred_h3_s2, "lightblue3")
draw_pred_ci(ages_monthly_s3, pred_h3_s3, "lightblue3")

  # fitted CI and median lines for all seasons
for (s in 1:(a3)) {
  ages_z_s <- season_start_ages[s] + z_fine * (season_end_ages[s] - season_start_ages[s])
  polygon(c(ages_z_s, rev(ages_z_s)),
          c(fitted_fine_lower[[s]], rev(fitted_fine_upper[[s]])),
          col = "grey70", border = NA)
  lines(ages_z_s, fitted_fine_median[[s]], col = "black", lwd = 1.5)
}

for (s in 1:a3) {
  ages_z_s <- season_start_ages[s] + z_fine * (season_end_ages[s] - season_start_ages[s])
  lines(ages_z_s, fitted_fine_median[[s]], col = "black", lwd = 1.5)
}

  # prediction median lines for 1,2,3 seasons ahead
draw_pred_med(ages_monthly_s1, pred_h3_s1)
draw_pred_med(ages_monthly_s2, pred_h3_s2) 
draw_pred_med(ages_monthly_s3, pred_h3_s3) 

  # observed performances
obs_train <- which(season.f[[athlete_idx]] <= (a1 - 1))
obs_test  <- which(season.f[[athlete_idx]] %in% c(a1, a2, a3))
obs_train_50m <- intersect(obs_train, pool_50m_idx)
obs_test_50m  <- intersect(obs_test,  pool_50m_idx)
points(exact_age[obs_train_50m], target.f[[athlete_idx]][obs_train_50m],
       pch = 19, cex = 0.6, col = "black")
points(exact_age[obs_test_50m], target.f[[athlete_idx]][obs_test_50m],
       pch = 1, cex = 0.6)

legend("bottomleft",
       legend = c("Training performances",
                  "Held-out performances",
                  "Fitted trajectory (median)",
                  "Fitted 95% CI",
                  "Predicted trajectory (median)",
                  "Predicted 95% CI"),
       col = c("black", "black", "black", "grey70", "black","lightblue3"),
       lty = c(NA, NA, 1, NA, 2, NA), pch = c(19, 1, NA, 15, NA, 15),
       lwd = c(NA, NA, 1.5, NA, 2, NA), bty = "n", cex = 0.6)
```

Log-Score for 1-6 Season Ahead Predictions (Horizons 1-6)
```{r}
#### Log-Score to Compare Uncertainties ####

# Helper function: Compute log-score
log_score <- function(pred_samples, actual) {
  # fit kernel density to posterior predictive samples
  dens <- density(pred_samples, n = 1024)
  
  # evaluate density at actual observed value
    # using linear interpolation between density points
  p_actual <- approx(dens$x, dens$y, xout = actual)$y
  
  # return negative log density (lower = better)
  -log(p_actual)
}

# Helper function: Prediction for one athlete, one horizon (similar to above)
predict_season_best <- function(fm.res, athlete_idx, target_season,
                                 horizon, n_samples, p_b, mean_age) {
  
  last_season_idx <- which(season.f[[athlete_idx]] == target_season)
  z_last <- times.f[[athlete_idx]][last_season_idx]
  ages_last <- ages.f[[athlete_idx]][last_season_idx]
  b_last <- TruncBernstein(z_last, degree = 4)
  
  pred_best <- rep(NA, n_samples)
  
  for (m in seq_len(n_samples)) {
    
    beta_a_m <- fm.res$beta_a[, m]
    sigmasq_inc_m <- fm.res$sigmasq_inc[m]
    sigmasq_rw3_m <- fm.res$sigmasq_rw_3[m]
    beta_bi_m <- fm.res$beta_bi[[m]][athlete_idx, ]
    sigmasqbeta_t_m <- fm.res$sigmasqbeta_t[, m]
    sigmasqbeta_t_ind_m <- fm.res$sigmasqbeta_t_ind[athlete_idx, m]
    
    # extend F forward from end of training seasons
    train_up_to  <- target_season - horizon
    F_obs <- fm.res$F[[athlete_idx]][1:train_up_to, m]
    F_ext <- extend_F(F_obs, n_future = horizon, sigmasq_inc  = sigmasq_inc_m,
                             sigmasq_rw_3 = sigmasq_rw3_m)
    eta_start <- F_ext[target_season - 1]
    eta_end <- F_ext[target_season]
    
    # draw within-season coefficients for target season
    beta_bit_new <- rnorm(p_b, mean = beta_bi_m, sd = sqrt(sigmasqbeta_t_ind_m * sigmasqbeta_t_m))
    
    # predict at each performance in target season
    preds <- rep(NA, length(last_season_idx))
    for (k in seq_along(last_season_idx)) {
      age_c <- ages_last[k] - mean_age
      des_mat <- c(1, age_c, age_c^2, age_c^3, age_c^4)
      g_a <- as.numeric(des_mat %*% beta_a_m[1:5])
      fi_z <- (1 - z_last[k]) * eta_start + z_last[k] * eta_end
      h_star <- as.numeric(b_last[k, ] %*% beta_bit_new)
      preds[k] <- g_a + fi_z + h_star
    }
    pred_best[m] <- min(preds)
  }
  return(pred_best)
}

# Helper function: Summary Statistics
print_results <- function(athlete_idx, actual, pred_list) {
  cat("\nSUMMARY STATISTICS: Athlete", athlete_idx, "\n")
  cat("Actual season best:", round(actual, 3), "seconds\n")
  cat(sprintf("%-12s %-10s %-10s %-10s %-10s %-10s\n",
            "Horizon", "Median", "Error", "CI Width", "In CI?", "Log Score"))
  for (h in seq_along(pred_list)) {
    pred <- pred_list[[h]]
    ls <- log_score(pred, actual)
    cat(sprintf("%-12s %-10f %-10f %-10f %-10s %-10f\n",
                paste(h, "season(s)"),
                median(pred),
                abs(median(pred) - actual),
                diff(quantile(pred, c(0.025, 0.975))),
                actual >= quantile(pred, 0.025) &
                actual <= quantile(pred, 0.975),
                ls))
  }
}


## ATHLETE 1
athlete_idx_a1 <- 1
horizons_a1 <- 6
n_seasons_a1 <- length(unique(season.f[[athlete_idx_a1]]))
target_season_a1 <- n_seasons_a1
actual_a1 <- min(target.f[[athlete_idx_a1]][season.f[[athlete_idx_a1]] == target_season_a1])

pred_a1 <- vector("list", horizons_a1)  # holder for each horizon's summary stats
for (h in 1:horizons_a1) {
  pred_a1[[h]] <- predict_season_best(
    fm.res, athlete_idx_a1, target_season_a1,
    horizon = h,
    n_samples = n_samples,
    p_b = p_b,
    mean_age = mean_age)
}

print_results(athlete_idx_a1, actual_a1, pred_a1)

## ATHLETE 51
athlete_idx_a51 <- 51
horizons_a51 <- 6
n_seasons_a51 <- length(unique(season.f[[athlete_idx_a51]]))
target_season_a51 <- n_seasons_a51
actual_a51 <- min(target.f[[athlete_idx_a51]][season.f[[athlete_idx_a51]] == target_season_a51])

pred_a51 <- vector("list", horizons_a51)
for (h in 1:horizons_a51) {
  pred_a51[[h]] <- predict_season_best(
    fm.res, athlete_idx_a51, target_season_a51,
    horizon = h,
    n_samples = n_samples,
    p_b = p_b,
    mean_age = mean_age)
}

print_results(athlete_idx_a51, actual_a51, pred_a51)


## THOUGHT: Is the lower log-score for Athlete 51 due to no gap years 
  # OR due to larger volume of data
# Use only first 9 seasons of Athlete 51 to match Athlete 1's data volume
target_season_a51_9 <- 9
horizons_a51_9 <- 6
actual_a51_9 <- min(target.f[[51]][season.f[[51]] == 9])

pred_a51_9 <- vector("list", horizons_a51_9)
for (h in 1:horizons_a51_9) {
  pred_a51_9[[h]] <- predict_season_best(
    fm.res, 51, target_season_a51_9,
    horizon = h, n_samples = n_samples,
    p_b = p_b, mean_age = mean_age)
}

cat("\nAthlete 51 - ONLY 9 Seasons")
print_results(athlete_idx_a51, actual_a51_9, pred_a51_9)
```

Log-Score for Layers 1,2,3
```{r}
#### Going Back to the Layer 1,2,3 Models ####
## Check Log-Score For Each of These ##
log_score_ext <- function(pred_samples, actual) {
  # extend range by 1 second in each direction to avoid NA at extremes
  dens <- density(pred_samples, n = 1024,
                      from = min(pred_samples) - 1,
                      to = max(pred_samples) + 1)
  p_actual <- approx(dens$x, dens$y, xout = actual)$y
  
  if (is.na(p_actual) || p_actual <= 0) { # Check edge cases 
    return(Inf)   # actual is in essentially a zero-probability region 
  }                  # so don't compute log-score 
  -log(p_actual)
}

ls_l1 <- log_score_ext(pred_season_best_l1, actual_season_best)
ls_l2 <- log_score_ext(pred_season_best_l2, actual_season_best)
ls_l3 <- log_score_ext(pred_season_best_l3, actual_season_best)

cat("Log scores by layer:\n")
cat("Layer 1:", round(ls_l1, 3), "\n")
cat("Layer 2:", round(ls_l2, 3), "\n")
cat("Layer 3:", round(ls_l3, 3), "\n")

# Check for which athlete these log-scores belong to
cat("Actual season best:", actual_season_best, "\n")
cat("Layer 1 range:", round(range(pred_season_best_l1), 3), "\n")
cat("Layer 2 range:", round(range(pred_season_best_l2), 3), "\n")
cat("Layer 3 range:", round(range(pred_season_best_l3), 3), "\n")

```

Multiple Athletes, Multiple Seasons Ahead Predictions
```{r}
# First find athletes with sufficient amount of data
  # check season counts for all athletes
season_counts.f <- sapply(1:100, function(i) length(unique(season.f[[i]])))
table(season_counts.f)
summary(season_counts.f)

  # which athletes have sufficient seasons for each horizon group
min_seasons_h3 <- 5    # to predict horizons 1-3
min_seasons_h6 <- 8    # to predict horizons 1-6

athletes_h3.f <- which(season_counts.f >= min_seasons_h3)
athletes_h6.f <- which(season_counts.f >= min_seasons_h6)

cat("Women athletes with >= 5 seasons (horizons 1-3):", length(athletes_h3.f), "\n")
cat("Women athlete indices:", athletes_h3.f, "\n\n")

cat("Women athletes with >= 8 seasons (horizons 1-6):", length(athletes_h6.f), "\n")
cat("Women athlete indices:", athletes_h6.f, "\n\n")

# Summary of how many athletes are excluded
cat("Excluded from h1-3 analysis:", 100 - length(athletes_h3.f), "athletes\n")
cat("Excluded from h1-6 analysis:", 100 - length(athletes_h6.f), "athletes\n")


# Storage for results
  # rows = athletes, cols = horizons
results_median.f <- matrix(NA, nrow = 100, ncol = 6)
results_error.f <- matrix(NA, nrow = 100, ncol = 6)
results_ci_width.f <- matrix(NA, nrow = 100, ncol = 6)
results_coverage.f <- matrix(NA, nrow = 100, ncol = 6)
results_logscore.f <- matrix(NA, nrow = 100, ncol = 6)
actual_bests.f <- rep(NA, 100)
n_seasons_all.f <- rep(NA, 100)


for (i in 1:100) {

  n_s <- season_counts.f[i]
  n_seasons_all.f[i] <- n_s
  
  # skip athletes with fewer than 5 seasons
  if (n_s < 5) {
    cat("(W) Athlete", i, ": skipped (only", n_s, "seasons)\n")
    next
  }
  
  cat("(W) Athlete", i, "(", n_s, "seasons)\n")
  
  # target season = last season
  target_s <- n_s
  last_idx <- which(season.f[[i]] == target_s)
  actual_best.f <- min(target.f[[i]][last_idx])
  actual_bests.f[i] <- actual_best.f
  
  # determine max horizon for this athlete
  max_h <- ifelse(n_s >= 8, 6, 3)
  
  # run predictions for each horizon
  for (h in 1:max_h) {
    
    pred <- predict_season_best(
      fm.res = fm.res,
      athlete_idx = i,
      target_season = target_s,
      horizon = h,
      n_samples = n_samples,
      p_b = p_b,
      mean_age = mean_age)
    
    results_median.f[i, h] <- median(pred)
    results_error.f[i, h] <- abs(median(pred) - actual_best.f)
    results_ci_width.f[i, h] <- diff(quantile(pred, c(0.025, 0.975)))
    results_coverage.f[i, h] <- actual_best.f >= quantile(pred, 0.025) &
                               actual_best.f <= quantile(pred, 0.975)
    results_logscore.f[i, h] <- log_score(pred, actual_best.f)
  }
}


# Summary Statistics
cat("\n SUMMARY: Horizons 1-3 (",length(athletes_h3.f), "(W) Athletes Included) \n")
cat(sprintf("%-15s %-10s %-10s %-10s %-10s\n",
            "Horizon", "Mean Error", "Coverage", "Mean CIW", "Mean LS"))

for (h in 1:3) {
  # only include athletes who had this horizon computed
  valid <- !is.na(results_error.f[, h])
  cat(sprintf("%-15s %-10f %-10f %-10f %-10f\n",
              paste(h, "season(s)"),
              mean(results_error.f[valid, h]),
              mean(results_coverage.f[valid, h]),
              mean(results_ci_width.f[valid, h]),
              mean(results_logscore.f[valid, h], na.rm = TRUE)))
}

cat("\nSUMMARY: Horizons 4-6 (", length(athletes_h6.f), "(W) Athletes Included) \n")
cat(sprintf("%-15s %-10s %-10s %-10s %-10s\n",
            "Horizon", "Mean Error", "Coverage", "Mean CIW", "Mean LS"))

for (h in 4:6) {
  valid.f <- !is.na(results_error.f[, h])
  cat(sprintf("%-15s %-10f %-10f %-10f %-10s\n",
              paste(h, "season(s)"),
              mean(results_error.f[valid.f, h]),
              mean(results_coverage.f[valid.f, h]),
              mean(results_ci_width.f[valid.f, h]),
              mean(results_logscore.f[valid.f, h], na.rm = TRUE)))
}


# Identify the Interesting Cases

# Best predicted athletes at horizon 1
best_h1.f <- order(results_error.f[, 1], na.last = TRUE)[1:5]
cat("\nBest Predicted Women's Athletes at Horizon 1:\n")
for (i in best_h1.f) {
  if (!is.na(results_error.f[i, 1]))
    cat("  Women's Athlete", i, ": error =", round(results_error.f[i, 1], 3),
        "log score =", round(results_logscore.f[i, 1], 3),
        "seasons =", n_seasons_all.f[i], "\n")
}

# Worst predicted athletes at horizon 1
worst_h1.f <- order(results_error.f[, 1], na.last = TRUE, decreasing = TRUE)[1:5]
cat("\nWorst Predicted Women's Athletes at Horizon 1:\n")
for (i in worst_h1.f) {
  if (!is.na(results_error.f[i, 1]))
    cat("  Women's Athlete", i, ": error =", round(results_error.f[i, 1], 3),
        "log score =", round(results_logscore.f[i, 1], 3),
        "seasons =", n_seasons_all.f[i], "\n")
}

# Athletes where log score is Inf (prediction completely missed)
inf_ls.f <- which(is.infinite(results_logscore.f[, 1]))
cat("\nWomen's athletes where Layer 3 prediction completely missed at horizon 1:\n")
cat("  Women's athlete indices:", inf_ls.f, "\n")
cat("  Number:", length(inf_ls.f), "out of", length(athletes_h3.f), "\n")


#  Save Results
validation_results.f <- list(
  results_median.f = results_median.f,
  results_error.f = results_error.f,
  results_ci_width.f = results_ci_width.f,
  results_coverage.f = results_coverage.f,
  results_logscore.f = results_logscore.f,
  actual_bests.f = actual_bests.f,
  n_seasons_all.f = n_seasons_all.f,
  athletes_h3.f = athletes_h3.f,
  athletes_h6.f = athletes_h6.f
)

saveRDS(validation_results.f, "validation_results.rds")


#results_median.f <- validation_results$results_median
#results_error.f <- validation_results$results_error
#results_ci_width.f <- validation_results$results_ci_width
#results_coverage.f <- validation_results$results_coverage
#results_logscore.f <- validation_results$results_logscore
#actual_bests.f <- validation_results$actual_bests
#n_seasons_all.f <- validation_results$n_seasons_all
#athletes_h3.f <- validation_results$athletes_h3
#athletes_h6.f <- validation_results$athletes_h6

# Clean infinite log scores
results_logscore_clean.f <- results_logscore.f
results_logscore_clean.f[is.infinite(results_logscore_clean.f)] <- NA


## THOUGHT: which affects predictions more - # of seasons or if there are gap years

# Check if number of seasons correlates with prediction accuracy
cor(n_seasons_all.f[athletes_h3.f], results_error.f[athletes_h3.f, 1], use = "complete.obs")

# Check if gap years correlate with poor predictions
  # identify athletes with gap years
has_gap_f <- sapply(1:100, function(i) {
  yrs <- sort(unique(f.swim.data$Year[f.swim.data$athlete == i]))
  any(diff(yrs) > 1)
})

  # compare mean error for gap vs no-gap year athletes at horizon 1
gap_athletes.f <- intersect(which(has_gap_f), athletes_h3.f)
no_gap_athletes.f <- intersect(which(!has_gap_f), athletes_h3.f)

cat("Gap year athletes (W) - mean error at H1:", 
    round(mean(results_error.f[gap_athletes.f, 1], na.rm = TRUE), 3), "\n")
cat("No gap year athletes (W) - mean error at H1:", 
    round(mean(results_error.f[no_gap_athletes.f, 1], na.rm = TRUE), 3), "\n")
cat("Gap year athletes (W) - mean log score at H1:", 
    round(mean(results_logscore_clean.f[gap_athletes.f, 1], na.rm = TRUE), 3), "\n")
cat("No gap year athletes (W) - mean log score at H1:", 
    round(mean(results_logscore_clean.f[no_gap_athletes.f, 1], na.rm = TRUE), 3), "\n")

# Which of the worst predicted athletes had gap years
cat("Gap year status of worst predicted women's athletes:\n")
for (i in c(20, 27, 76, 22, 63)) {
  yrs <- sort(unique(f.swim.data$Year[f.swim.data$athlete == i]))
  cat("(W) Athlete", i, ":", n_seasons_all.f[i], "seasons,",
      "gap year:", any(diff(yrs) > 1), "\n")
}

  # investigate worst predicted non-gap year athletes
for (i in c(27, 22, 63)) {
  yrs <- sort(unique(f.swim.data$Year[f.swim.data$athlete == i]))
  n_s <- n_seasons_all.f[i]
  last_s <- n_s
  last_idx <- which(season.f[[i]] == last_s)
  
  cat("\n(W) Athlete", i, ":\n")
  cat("  Seasons:", n_s, "| Years:", min(yrs), "to", max(yrs), "\n")
  cat("  Actual season best:", round(actual_bests.f[i], 3), "seconds\n")
  cat("  Predicted median at H1:", round(results_median.f[i, 1], 3), "seconds\n")
  cat("  Error at H1:", round(results_error.f[i, 1], 3), "seconds\n")
  cat("  Log score at H1:", round(results_logscore.f[i, 1], 3), "\n")
  
  # print all season bests to see career trajectory
  cat("  Season bests:\n")
  for (s in 1:n_s) {
    s_idx <- which(season.f[[i]] == s)
    cat("    Season", s, ":", round(min(target.f[[i]][s_idx]), 3), 
        "seconds (", length(s_idx), "performances)\n")
  }
}

# Which of the best predicted athletes had gap years
cat("Gap year status of best predicted women's athletes:\n")
for (i in c(83, 72, 96, 97, 49)) {
  yrs <- sort(unique(f.swim.data$Year[f.swim.data$athlete == i]))
  cat("(W) Athlete", i, ":", n_seasons_all.f[i], "seasons,",
      "gap year:", any(diff(yrs) > 1), "\n")
}

for (i in c(83, 72, 96, 97, 49)) {
  yrs <- sort(unique(f.swim.data$Year[f.swim.data$athlete == i]))
  n_s <- n_seasons_all.f[i]
  
  cat("\n(W) Athlete", i, ":\n")
  cat("  Seasons:", n_s, "| Years:", min(yrs), "to", max(yrs), "\n")
  cat("  Gap year:", any(diff(yrs) > 1), "\n")
  cat("  Actual season best:", round(actual_bests.f[i], 3), "seconds\n")
  cat("  Predicted median H1:", round(results_median.f[i, 1], 3), "seconds\n")
  cat("  Error H1:", round(results_error.f[i, 1], 3), "seconds\n")
  cat("  Log score H1:", round(results_logscore.f[i, 1], 3), "\n")
  
  cat("  Season bests:\n")
  for (s in 1:n_s) {
    s_idx <- which(season.f[[i]] == s)
    cat("    Season", s, ":", round(min(target.f[[i]][s_idx]), 3),
        "seconds (", length(s_idx), "performances)\n")
  }
}



early_gap_f <- c()
late_gap_f  <- c()

for (i in which(has_gap_f)) {
  yrs <- sort(unique(f.swim.data$Year[f.swim.data$athlete == i]))
  gap_positions <- which(diff(yrs) > 1)
  n_s <- length(unique(season.f[[i]]))
  
  if (any(gap_positions >= (n_s - 3))) {
    late_gap_f <- c(late_gap_f, i)
  } else {
    early_gap_f <- c(early_gap_f, i)
  }
}

cat("Female athletes with early career gap:", length(early_gap_f), "\n")
cat("Female athletes with late career gap:", length(late_gap_f), "\n")

cat("Early gap mean log score H1:", 
    round(mean(results_logscore_clean.f[early_gap_f, 1], na.rm=TRUE), 3), "\n")
cat("Late gap mean log score H1:", 
    round(mean(results_logscore_clean.f[late_gap_f, 1], na.rm=TRUE), 3), "\n")


cat("Female early gap in h3 analysis:", 
    length(intersect(early_gap_f, athletes_h3.f)), "\n")
cat("Female late gap in h3 analysis:", 
    length(intersect(late_gap_f, athletes_h3.f)), "\n")
```

Log-Score Degradation Plots
```{r} 
# First create naive baseline for comparison
  # naive baseline: predict using historical (career) mean and sd of season bests
naive_log_score <- function(athlete_idx, target_season) {
  
  # compute historical season bests for all seasons except the target
  historical_bests <- sapply(1:(target_season - 1), function(s) {
    s_idx <- which(season.f[[athlete_idx]] == s)
    min(target.f[[athlete_idx]][s_idx])
  })
  
  # fit a normal distribution to historical season bests
  hist_mean <- mean(historical_bests)
  hist_sd <- sd(historical_bests)
  
  # actual season best
  actual <- min(target.f[[athlete_idx]][season.f[[athlete_idx]] == target_season])
  
  # log score of normal distribution at actual value
  -log(dnorm(actual, mean = hist_mean, sd = hist_sd))
}

naive_baselines.f <- rep(NA, 100)
for (i in athletes_h3.f) {
  naive_baselines.f[i] <- naive_log_score(i, n_seasons_all.f[i])
}
cat("Female mean naive baseline:", 
    round(mean(naive_baselines.f[athletes_h3.f], na.rm=TRUE), 3), "\n")

  # compute mean log scores across all athletes at each horizon (1-6)
    # overall, gap, and no-gap separately
mean_ls.f <- rep(NA, 6)
mean_ls_gap.f <- rep(NA, 6)
mean_ls_no_gap.f <- rep(NA, 6)
sd_ls.f <- rep(NA, 6)   # standard deviation for error bars

for (h in 1:6) {
  # all athletes
  valid.f <- athletes_h3.f[!is.na(results_logscore_clean.f[athletes_h3.f, h])]
  if (h > 3) {
    valid <- athletes_h6.f[!is.na(results_logscore_clean.f[athletes_h6.f, h])]
  }
  mean_ls.f[h] <- mean(results_logscore_clean.f[valid.f, h])
  sd_ls.f[h] <- sd(results_logscore_clean.f[valid.f, h])
  
  # gap athletes
  gap_valid.f <- gap_athletes.f[!is.na(results_logscore_clean.f[gap_athletes.f, h])]
  mean_ls_gap.f[h] <- mean(results_logscore_clean.f[gap_valid.f, h])
  
  # no-gap athletes
  no_gap_valid.f <- no_gap_athletes.f[!is.na(results_logscore_clean.f[no_gap_athletes.f, h])]
  mean_ls_no_gap.f[h] <- mean(results_logscore_clean.f[no_gap_valid.f, h])
}

  # mean naive baseline
mean_naive.f <- mean(naive_baselines.f[athletes_h3.f], na.rm = TRUE)

# Look at how log-score degrade as horizons increases
cat("Mean Log-Scores By Horizon:\n")
cat(sprintf("%-15s %-8s %-8s %-8s\n", "Horizon", "All", "Gap", "No Gap"))
for (h in 1:6) {
  cat(sprintf("%-15s %-8.4f %-8.4f %-8.4f\n",
              paste(h, "season(s)"),
              mean_ls.f[h],
              mean_ls_gap.f[h],
              mean_ls_no_gap.f[h]))
}
cat(sprintf("%-15s %-8.4f\n", "Naive baseline", mean_naive.f))

# Plot (want to see if there is a clear "elbow")
plot(NA, NA,
     xlim = c(0.8, 6.2),
     ylim = range(c(mean_ls_gap.f, mean_ls_no_gap.f, mean_naive.f), 
                  na.rm = TRUE) + c(-0.2, 0.2),
     xlab = "Forecast Horizon (Seasons Ahead)",
     ylab = "Mean Log Score (lower = better)",
     main = "(W) Log Score Degradation by Career Continuity",
     xaxt = "n")
axis(1, at = 1:6)

  # naive baseline reference line
abline(h = mean_naive.f, col = "grey", lty = 3, lwd = 1.5)

  # all athletes
lines(1:6, mean_ls.f, type = "b", pch = 19, col = "black", lwd = 2, cex = 1.2)

  # gap year athletes
lines(1:6, mean_ls_gap.f, type = "b", pch = 17, col = "tomato", lwd = 2, lty = 2, cex = 1.2)

  # no gap athletes
lines(1:6, mean_ls_no_gap.f, type = "b", pch = 15, col = "darkseagreen", lwd = 2, lty = 2, cex = 1.2)

  # add error bars (±1 SD) for all athletes line
arrows(1:6, mean_ls.f - sd_ls.f, 1:6, mean_ls.f + sd_ls.f,
       angle = 90, code = 3, length = 0.05,
       col = "grey50", lwd = 1)

legend("topleft",
       legend = c("All Athletes (W)",
                  "Gap Year Athletes",
                  "No Gap Athletes",
                  "Error Bars (±1 SD)",
                  "Naive Baseline"),
       col    = c("black", "tomato", "darkseagreen", "grey50", "grey"),
       lty    = c(1, 2, 2, 1, 3),
       pch    = c(19, 17, 15, NA, NA),
       lwd    = 2, bty = "n", cex = 0.9)
```

CALIBRATION
```{r}
# First need to rerun the prediction (same as before) 
  # but save the raw samples (not just median & ci)

pred_samples_all_f <- array(NA, dim = c(n_samples, 100, 6))

for (i in athletes_h3.f) {
  max_h.f <- ifelse(i %in% athletes_h6.f, 6, 3)
  for (h in 1:max_h.f) {
    pred_samples_all_f[, i, h] <- predict_season_best(
      fm.res = fm.res,
      athlete_idx = i,
      target_season = n_seasons_all.f[i],
      horizon = h,
      n_samples = n_samples,
      p_b = p_b,
      mean_age = mean_age)
  }
}

#saveRDS(pred_samples_all_f, "pred_samples_all_f.rds")


# Calibration calculation
alpha_levels <- c(0.50, 0.60, 0.70, 0.80, 0.90, 0.95)
coverage_mat_f <- matrix(NA, nrow = 6, ncol = length(alpha_levels))  # Holder

  # calculate how many of the actual season bests lie in the ci
for (h in 1:6) {
  eligible <- if (h <= 3) athletes_h3.f else athletes_h6.f
  for (a_idx in seq_along(alpha_levels)) {
    alpha <- alpha_levels[a_idx]  # nominal coverage level
    lower_q <- (1 - alpha) / 2    # ci lower bound
    upper_q <- 1 - lower_q        # ci upper bound
    
    in_ci <- rep(NA, 100)
    for (i in eligible) {
      pred <- pred_samples_all_f[, i, h]
      actual <- actual_bests.f[i]
      # TRUE if actual season best falls within the alpha% CI
      in_ci[i] <- actual >= quantile(pred, lower_q, na.rm = TRUE) &
                  actual <= quantile(pred, upper_q, na.rm = TRUE)
    }
    # observed coverage = proportion of athletes where actual falls in CI
    coverage_mat_f[h, a_idx] <- mean(in_ci, na.rm = TRUE)
  }
}

# Coverage table
cat("(W) Calibration Results (By Horizon):\n")
cat(sprintf("%-10s %-8s %-8s %-8s %-8s %-8s %-8s\n", 
            "Horizon", "50%", "60%", "70%", "80%", "90%", "95%"))
for (h in 1:6) {
  cat(sprintf("%-10s %-8.4f %-8.4f %-8.4f %-8.4f %-8.4f %-8.4f\n",
              paste("H", h),
              coverage_mat_f[h, 1],
              coverage_mat_f[h, 2],
              coverage_mat_f[h, 3],
              coverage_mat_f[h, 4],
              coverage_mat_f[h, 5],
              coverage_mat_f[h, 6]))
}

# Plot
plot(NA, NA,
     xlim = c(0.25, 1), ylim = c(0.25, 1),
     xlab = "Nominal Coverage", ylab = "Observed Coverage",
     main = "(W) Calibration Plot by Prediction Horizon")

  # perfect calibration = diagonal line
abline(0, 1, col = "grey50", lty = 2, lwd = 1.5)

  # add reference lines
abline(h = seq(0.3, 1, by = 0.1), col = "grey90", lty = 1)
abline(v = seq(0.3, 1, by = 0.1), col = "grey90", lty = 1)

  # plot horizons 1,3,5
cols <- c("steelblue", "mediumpurple", "goldenrod")

lines(alpha_levels, coverage_mat_f[1, ], type = "b", pch = 19,
        col = cols[1], lwd = 2, cex = 1.2)
lines(alpha_levels, coverage_mat_f[3, ], type = "b", pch = 19,
        col = cols[2], lwd = 2, cex = 1.2)
lines(alpha_levels, coverage_mat_f[5, ], type = "b", pch = 19,
        col = cols[3], lwd = 2, cex = 1.2)

legend("topleft",
       legend = c("Perfect calibration",
                  "Horizon 1", "Horizon 3", "Horizon 5"),
       col    = c("grey50", cols),
       lty    = c(2, 1, 1, 1),
       pch    = c(NA, 19, 19, 19),
       lwd    = 2, bty = "n")

```

Calibration by Gap Year/No-Gap Year
```{r}
# Holders
coverage_gap.f <- matrix(NA, nrow = 3, ncol = length(alpha_levels))
coverage_no_gap.f <- matrix(NA, nrow = 3, ncol = length(alpha_levels))

# Redefine gap athletes
has_gap.f <- sapply(1:100, function(i) {
  yrs <- sort(unique(f.swim.data$Year[f.swim.data$athlete == i]))
  any(diff(yrs) > 1)
})

# Separate gap athletes from no-gap athletes
gap_athletes_h3.f <- intersect(which(has_gap.f),  athletes_h3.f)
no_gap_athletes_h3.f <- intersect(which(!has_gap.f), athletes_h3.f)

# Calibration computation (same as above)
for (h in 1:3) {
  for (a_idx in seq_along(alpha_levels)) {
    alpha <- alpha_levels[a_idx]
    lower_q <- (1 - alpha) / 2
    upper_q <- 1 - lower_q
    
    in_ci_gap <- rep(NA, 100)
    in_ci_no_gap <- rep(NA, 100)
    
    for (i in gap_athletes_h3.f) {
      pred <- pred_samples_all_f[, i, h]
      in_ci_gap[i] <- actual_bests.f[i] >= quantile(pred, lower_q, na.rm = TRUE) &
                      actual_bests.f[i] <= quantile(pred, upper_q, na.rm = TRUE)
    }
    for (i in no_gap_athletes_h3.f) {
      pred <- pred_samples_all_f[, i, h]
      in_ci_no_gap[i] <- actual_bests.f[i] >= quantile(pred, lower_q, na.rm = TRUE) &
                         actual_bests.f[i] <= quantile(pred, upper_q, na.rm = TRUE)
    }
    
    coverage_gap.f[h, a_idx] <- mean(in_ci_gap, na.rm = TRUE)
    coverage_no_gap.f[h, a_idx] <- mean(in_ci_no_gap, na.rm = TRUE)
  }
}

# Coverage table
cat("(W) Calibration by Career Continuity (Horizon 1):\n")
cat(sprintf("%-12s %8s %8s %8s %8s %8s %8s\n",
            "", "50%", "60%", "70%", "80%", "90%", "95%"))
cat(sprintf("%-12s %8.3f %8.3f %8.3f %8.3f %8.3f %8.3f\n",
            "Gap",
            coverage_gap.f[1,1], coverage_gap.f[1,2], coverage_gap.f[1,3],
            coverage_gap.f[1,4], coverage_gap.f[1,5], coverage_gap.f[1,6]))
cat(sprintf("%-12s %8.3f %8.3f %8.3f %8.3f %8.3f %8.3f\n",
            "No Gap",
            coverage_no_gap.f[1,1], coverage_no_gap.f[1,2], coverage_no_gap.f[1,3],
            coverage_no_gap.f[1,4], coverage_no_gap.f[1,5], coverage_no_gap.f[1,6]))

# Calibration Plot
  # by gap status for horizon 1
plot(NA, NA,
     xlim = c(0.2, 1), ylim = c(0.2, 1),
     xlab = "Nominal Coverage", ylab = "Observed Coverage",
     main = "(W) Calibration by Career Continuity (Horizon 1)")

  # perfect calibration
abline(0, 1, col = "grey50", lty = 2, lwd = 1.5)

  # reference lines
abline(h = seq(0.3, 1, by = 0.1), col = "grey90")
abline(v = seq(0.3, 1, by = 0.1), col = "grey90")

  # plot gap year & no-gap year calibrations 
lines(alpha_levels, coverage_gap.f[1, ], type = "b", pch = 19, col = "tomato", lwd = 2)
lines(alpha_levels, coverage_no_gap.f[1, ], type = "b", pch = 19, col = "darkseagreen", lwd = 2)

legend("topleft",
       legend = c("Perfect Calibration", 
                  "Gap Year Athletes",
                  "No Gap Athletes"),
       col = c("grey50", "tomato", "darkseagreen"),
       lty = c(2, 1, 1), pch = c(NA, 19, 19),
       lwd = 2, bty = "n")
```

Calibration Plot for Layers 1,2,3
```{r}
# Holder
pred_samples_layers.f <- array(NA, dim = c(n_samples, 100, 3))

# Rerun Layers 1,2,3 predictions for all athletes (same as before)
for (i in athletes_h3.f) {
  
  last_s <- n_seasons_all.f[i]
  last_idx <- which(season.f[[i]] == last_s)
  z_last_i <- times.f[[i]][last_idx]
  ages_last_i <- ages.f[[i]][last_idx]
  b_last_i <- TruncBernstein(z_last_i, degree = 4)
  
  pred_l1 <- rep(NA, n_samples)
  pred_l2 <- rep(NA, n_samples)
  pred_l3 <- rep(NA, n_samples)
  
  for (m in seq_len(n_samples)) {
    
    beta_a_m <- fm.res$beta_a[, m]
    sigmasq_inc_m <- fm.res$sigmasq_inc[m]
    sigmasq_rw3_m <- fm.res$sigmasq_rw_3[m]
    beta_bi_m <- fm.res$beta_bi[[m]][i, ]
    sigmasqbeta_t_m <- fm.res$sigmasqbeta_t[, m]
    sigmasqbeta_t_ind_m <- fm.res$sigmasqbeta_t_ind[i, m]
    
    # extend F by 1 season (horizon 1 for all layers)
    F_obs <- fm.res$F[[i]][1:(last_s - 1), m]
    F_ext <- extend_F(F_obs, n_future = 1, sigmasq_inc = sigmasq_inc_m,
                          sigmasq_rw_3 = sigmasq_rw3_m)
    eta_start <- F_ext[last_s - 1]
    eta_end <- F_ext[last_s]

    beta_bit_new <- rnorm(p_b, mean = beta_bi_m, sd = sqrt(sigmasqbeta_t_ind_m * sigmasqbeta_t_m))
    
    preds_l1 <- rep(NA, length(last_idx))
    preds_l2 <- rep(NA, length(last_idx))
    preds_l3 <- rep(NA, length(last_idx))
    
    for (k in seq_along(last_idx)) {
      age_c <- ages_last_i[k] - mean_age
      des_mat <- c(1, age_c, age_c^2, age_c^3, age_c^4)
      g_a <- as.numeric(des_mat %*% beta_a_m[1:5])
      fi_z <- (1 - z_last_i[k]) * eta_start + z_last_i[k] * eta_end
      h_star <- as.numeric(b_last_i[k, ] %*% beta_bit_new)
      
      preds_l1[k] <- g_a                    # Layer 1: g(a) only
      preds_l2[k] <- g_a + fi_z             # Layer 2: g(a) + f_i(z)
      preds_l3[k] <- g_a + fi_z + h_star    # Layer 3: full model
    }
    
    pred_l1[m] <- min(preds_l1)
    pred_l2[m] <- min(preds_l2)
    pred_l3[m] <- min(preds_l3)
  }
  
  pred_samples_layers.f[, i, 1] <- pred_l1
  pred_samples_layers.f[, i, 2] <- pred_l2
  pred_samples_layers.f[, i, 3] <- pred_l3
}

saveRDS(pred_samples_layers.f, "pred_samples_layers.rds")

# Calibration Computation (same as before) 
alpha_levels  <- c(0.50, 0.60, 0.70, 0.80, 0.90, 0.95)
coverage_layers.f <- matrix(NA, nrow = 3, ncol = length(alpha_levels))

for (l in 1:3) {
  for (a_idx in seq_along(alpha_levels)) {
    alpha <- alpha_levels[a_idx]
    lower_q <- (1 - alpha) / 2
    upper_q <- 1 - lower_q
    
    in_ci <- rep(NA, 100)
    for (i in athletes_h3.f) {
      pred <- pred_samples_layers.f[, i, l]
      actual   <- actual_bests.f[i]
      in_ci[i] <- actual >= quantile(pred, lower_q, na.rm = TRUE) &
                  actual <= quantile(pred, upper_q, na.rm = TRUE)
    }
    coverage_layers.f[l, a_idx] <- mean(in_ci, na.rm = TRUE)
  }
}

# Calibration Table
cat("(W) Calibration Results (By Layer):\n")
cat(sprintf("%-10s %-8s %-8s %-8s %-8s %-8s %-8s\n", 
            "Horizon", "50%", "60%", "70%", "80%", "90%", "95%"))
for (l in 1:3) {
  cat(sprintf("%-10s %-8.4f %-8.4f %-8.4f %-8.4f %-8.4f %-8.4f\n",
              paste("Layer", l),
              coverage_layers.f[l,1],
              coverage_layers.f[l,2],
              coverage_layers.f[l,3],
              coverage_layers.f[l,4],
              coverage_layers.f[l,5],
              coverage_layers.f[l,6]))
}


# Plot
plot(NA, NA,
     xlim = c(0.45, 1), ylim = c(0, 1),
     xlab = "Nominal Coverage", ylab = "Observed Coverage",
     main = "(W) Calibration Plot by Prediction Layer (Horizon 1)")

abline(0, 1, col = "grey50", lty = 2, lwd = 1.5)
abline(h = seq(0, 1, by = 0.1), col = "grey90")
abline(v = seq(0.5, 1, by = 0.1), col = "grey90")

cols <- c("plum", "seagreen", "steelblue")
for (l in 1:3) {
  lines(alpha_levels, coverage_layers.f[l, ],
        type = "b", pch = 19,
        col = cols[l], lwd = 2, cex = 1.2)
}

legend("topleft",
       legend = c("Perfect calibration",
                  "Layer 1: g(a) only",
                  "Layer 2: g(a) + f_i(z)",
                  "Layer 3: Full Model"),
       col = c("grey50", cols),
       lty = c(2, 1, 1, 1),
       pch = c(NA, 19, 19, 19),
       lwd = 2, bty = "n")

cat("Layer 3 female 95% coverage:", 
    round(coverage_layers.f[3, 6], 3), "\n")
cat("Layer 2 female 95% coverage:", 
    round(coverage_layers.f[2, 6], 3), "\n")
cat("Layer 1 female 95% coverage:", 
    round(coverage_layers.f[1, 6], 3), "\n")
```

